Missing the Point: Non-Convergence in Iterative Imputation Algorithms

### 2020-06-04 ICML/ARTEMISS_proposal.tex
Fill in ICML format from thesis manuscript
Also change label x axis in plot from missingness proportion to proportion of missing cases, and some other things.

A popular method to accommodate missing data is to \textit{impute} (i.e., `fill in') the missing values in an incomplete dataset. It is widely accepted that imputation techniques such as multiple imputation (MI; @rubin76) can yield statistically valid inferences. The validity of these inferences relies on the method through which imputations are obtained---often iterative algorithms. Iterative imputation algorithms are employed in e.g., $\mathtt{SPSS}$, $\mathtt{Stata}$, and the $\mathtt{R}$ packages $\mathtt{MI}$, and $\mathtt{mice}$. \textbf{Or use:} Convergence of the algorithm used to solve the missing data problem is a topic that has not received much attention but is ever so important, as most imputation software packages draw inference from iterative imputation procedures.

The validity of the inference all depends on the state space of the algorithm at the final iteration. If the algorithm does not reach convergence, ``the imputed datasets will not be truly independent and the variability among them may understate the true levels of missing-data uncertainty'' (Schafer and Olsen, p. 556).

But when is this the case? Current practice of visually inspecting imputations is not sufficient because 1) no objective point at which convergence is established, 2) only severely pathological cases of non-convergence may be identified, and 3) diagnosing non-convergence can be challenging to the untrained eye. On top of that, the statistics to inspect are either univariate or dependent on the substantive model of interest. Converged univariate statistics do not guarantee multivariate convergence, and with a model-dependent statistic there is no guarantee that the algorithm is converged \textit{enough} for other models of scientific interest. This negates the advantages of MI: solving the missing data problem and complete data problem separately.

Therefore, we propose a novel parameter to inspect. We use this parameter to identify non-convergence with two quantitative diagnostic methods: Rhat and AC.

We simulate and investigate the plausibility of identifiers for non-convergence for iterative imputation algorithms ...

@@

Valid inferences may be obtained when there is still non-convergence in the algorithm.

@@

\bibliography{example_paper}
\bibliographystyle{icml2020}

>>>>

A popular method to accommodate missing data is to \textit{impute} (i.e., `fill in') the missing values in an incomplete dataset. It is widely accepted that imputation techniques such as multiple imputation \citep[MI;][]{rubin76} can yield statistically valid inferences. The validity of these inferences relies on the method through which imputations are obtained---often iterative algorithms. Iterative imputation algorithms are employed in e.g., $\mathtt{SPSS}$, $\mathtt{Stata}$, and the $\mathtt{R}$ packages $\mathtt{MI}$, and $\mathtt{mice}$. \textit{[Or use:} Convergence of the algorithm used to solve the missing data problem is a topic that has not received much attention but is ever so important, as most imputation software packages draw inference from iterative imputation procedures.]

With iterative imputation, the validity of the inference depends on the state-space of the algorithm at the final iteration. This introduces a potential threat to the validity of the imputations: What if the algorithm has not converged? Are the imputations then to be trusted? And can we rely on the inference obtained on the completed data?

These remain open questions since the convergence properties of iterative imputation algorithms have not been systematically studied \citep{buur18}.\footnote{If the algorithm does not reach convergence, ``the imputed datasets will not be truly independent and the variability among them may understate the true levels of missing-data uncertainty'' (Schafer and Olsen, p. 556).} There is no scientific consensus on how to evaluate the convergence of imputation algorithms \citep{zhu15, taka17}. Moreover, the behavior of such algorithms under certain default imputation models (e.g., `predictive mean matching') is an entirely open question \citep{murr18}. Therefore, algorithmic convergence should be monitored carefully.

The current practice of visually inspecting imputations may be undesirable for several reasons: 1) it may be challenging to the untrained eye, 2) only severely pathological cases of non-convergence may be diagnosed, and 3) there is not an objective measure that quantifies convergence \citep[\(\S\) 6.5.2]{buur18}. On top of that, the recommended parameters to inspect may be insufficient. The parameters are scalar summaries of the state-space of the algorithm. They are either univariate, or depend on the substantive model of interest. Apparent convergence of a univariate parameter does not guarantee convergence of the multivariate state-space of the algorithm. And with a model-dependent statistic there is no guarantee that the algorithm is converged \textit{enough} for other models of scientific interest. Additionally, model-dependency negates an important advantage of MI: solving the missing data problem separately from the scientific problem.

Therefore, we propose a novel parameter to inspect. We use this parameter to identify non-convergence with two quantitative diagnostic methods. We use model-based simulation to investigate the plausibility of identifiers for non-convergence for iterative imputation algorithms and provide guidelines for empirical research. For reasons of brevity, we only focus on the iterative imputation algorithm implemented in the popular \texttt{mice} package in \texttt{R} \citep{mice, R}. \textit{[Provide notation here? E.g., the number of imputations, iterations, etc.]}

\section{Methods}
\label{methods}

\subsection{Diagnostic Methods}
We consider several sets of diagnostic methods to identify non-convergence in the iterative imputation algorithm. Each method is a combination of a parameter that summarizes the state-space of the algorithm, $\theta$, and an identifier. As $\theta$s we use the default parameters of the \texttt{MICE} algorithm (i.e., the univariate $\theta$s chain means and chain variances), a model-dependent user-specified multivariate parameter $Q$, and a novel parameter: $\lambda_1$. $\lambda_{1}$ has the appealing property that it is not dependent on the model of scientific interest, yet still summarizes the multivariate state-space of the algorithm.

We define $\lambda_{1}$ as the first eigenvalue of the variance-covariance matrix of the completed data.\footnote{By definition, this is equal to the variance of the first component of the principal component analylis (PCA) solution on the same data.} For each imputation $\ell$, let $\lambda_{1, \ell} \geq \lambda_{2, \ell} \geq ... \geq \lambda_{j, \ell}$ be the eigenvalues of the variance-covariance matrix $S_{\ell}$, where $\ell = 1, ..., m$. Since the eigenvalues of a variance-covariance matrix summarize the total set of covariances in the data, $\lambda_1$ captures the multivariate nature of the completed data without imposing a substantive model of interest.

The identifiers are as recommended by e.g. \cite{cowl96}, namely the potential scale reduction factor $\widehat{R}$ and autocorrelation. $\widehat{R}$ is also known as the `Gelman-Rubin statistic' after the original paper \cite{gelm92}. We use a recently proposed adapted version of $\widehat{R}$ \citep{veht19}. We diagnose non-convergence if $\widehat{R}$-values exceed a certain diagnostic threshold. The thresholds under consideration are the traditional threshold $\widehat{R}=1.2$ \citep{gelm92}, the common $\widehat{R}=1.1$ \citep{gelm13}, and the recent $\widehat{R}=1.01$ \citep{veht19}.

We diagnose non-convergence with autocorrelation when the correlation between two subsequent $\theta$-values within the same chain is significantly different from zero \citep[p.~147, p.~42]{lync07, hynd18}.

\subsection{Simulation Study}
We investigate non-convergence of the iterative imputation algorithm implemented in the $\mathtt{R}$ package $\mathtt{mice}$ through model-based simulation. The simulation set-up is summarized in pseudo-code (see Algorithm \ref{alg:setup}). The complete script and technical details are available from \href{https://github.com/hanneoberman/MissingThePoint}{github.com/hanneoberman/MissingThePoint}. \textit{[Define conditions here somewhere, and $p_{\rm inc}$and $T$.]}

\begin{algorithm}
   \caption{Simulation set-up}
   \label{alg:setup}
\begin{algorithmic}
\STATE Simulate data 
\REPEAT 
\FOR {$p_{\rm inc} = .05, .25, .50, .75, .95$ (missingness conditions)}
  \STATE Create missingness
  \FOR {$T = 1, 2, ..., 100$ (early stopping conditions)}
   \STATE Impute missingness
   \STATE Perform analysis of scientific interest
   \STATE Compute non-convergence diagnostics 
   \STATE Pool results across imputations
   \STATE Compute performance measures
   \ENDFOR
 \ENDFOR 
 \STATE Combine outcomes of all conditions
\UNTIL{$n_{\rm sim} = 1000$ (simulation repetitions)}
\STATE Aggregate outcomes across simulation runs
\end{algorithmic}
\end{algorithm}

The aim of the simulation study is twofold: 1) to determine the effect of early stopping and missingness conditions on the validity of statistical inferences, 2) to investigate the plausibility of identifiers for non-convergence.

We assess the validity of statistical inferences with bias, coverage rate and confidence interval width as performance measures (see @buur18 for definitions). And we evaluate the plausibility of the identifiers using the diagnostic thresholds for $\widehat{R}$ and autocorrelation (i.e., $\widehat{R}=1.2,1.1,1.01$ and the critical value for AC = 0).

\textit{[Very incomplete from here!]} We consider several quantities of scientific interest.

The statistical inference of interest is multivariate linear regression.

The data generating mechanism of the predictor space is a multivariate normal distribution

\section{Results}
\label{results}

\begin{figure*}
{\centering \includegraphics{Figures/estimates} 
}
\caption{Impact of non-convergence on statistical inferences. Depicted are the bias, coverage rate (CR) and confidence interval width (CIW) of the worst-performing quantities of scientific interest $Q$ in terms of bias. The gray lines represent the objectives: unbiased estimates with nominal coverage.}\label{fig:est}
\end{figure*}

Results are displayed in Figure \ref{fig:est} and \ref{fig:diag}. Simulation conditions with a high proportion of missing cases and a low number of iterations most often result in biased estimates and non-nominal coverages.

\textit{[Copy-paste from here!]} These results demonstrate that the estimates of univariate scientific estimands \(Q\) are not impacted by early stopping of the MICE algorithm or the proportion of incomplete cases in \(y\). Unbiased estimates may be obtained after just one iteration. Multivariate estimates, by contrast, are affected by both the number of iterations and the proportion of missing cases. Completely unbiased estimates are only obtained under low to moderate missingness (\(p_{\rm inc}\leq.50\)), after at most three iterations. We observe approximately unbiased estimates after at most seven iterations (for any \(p_{\rm inc}\) considered). This implies that the algorithm produces stable, non-improving estimates when \(T\geq7\).

\begin{figure*}
{\centering \includegraphics{Figures/diagnostics} 
}
\caption{Non-convergence identified by diagnostic methods: $\widehat{R}$ and $AC$ applied on several $\theta$s. The left-hand side of the figure contains $\widehat{R}$-values, and the right-hand side contains $AC$-values. Depicted in the rows are the scalar summaries $\theta$: chain mean, chain variance, the quantity of scientific interest $\hat{Q}$, and the first eigenvalue of the variance-covariance matrix $\lambda_1$.}\label{fig:diag}
\end{figure*}

Overall, the methods to diagnose non-convergence perform as expected: they indicate more signs of non-convergence in conditions with worse performance in terms of bias and confidence-validity. Under low to moderate missingness, it does not seem to matter on which \(\theta\) the non-convergence diagnostics are applied. If we look at complete unbiasedness, however, we notice that univariate \(\theta\)s fail to diagnose the persistent bias in conditions where \(p_{\rm inc}\geq.75\).

The number of iterations necessary to obtain approximately unbiased, confidence-valid estimates corresponds to the \(\widehat{R}\) threshold 1.2. The thresholds 1.1 and 1.01 seem too strict compared to the validity of the inferences. The threshold to diagnose significant \(AC\)-values does not appear to be appropriate in at least the current set-up, and perhaps in iterative imputation in general. A better heuristic to diagnose non-stationarity with \(AC\) may be through evaluation of the \(AC\)-values across iterations: If the values do not substantially decrease with \(T\), approximate stationarity may be concluded.

@@

We have shown that iterative imputation algorithms can yield correct outcomes, even when a converged state has not yet formally been reached. Any further iterations would then burn computational resources without improving the statistical inferences. Our study found that---in the cases considered---inferential validity was achieved after five to ten iterations, much earlier than indicated by the \(\widehat{R}\) and AC diagnostics. Of course, it never hurts to iterate longer, but such calculations hardly bring added value.

@@

\bibliography{missing_the_point}
\bibliographystyle{icml2020}

### 2020-06-04 Writeup/2.Manuscript.Rmd
Fill in ICML format from thesis manuscript
Also change label x axis in plot from missingness proportion to proportion of missing cases, and some other things.

```{r Qs, fig.height = 6.8, fig.cap="Impact of non-convergence on statistical inferences. Depicted are the bias, coverage rate (CR) and confidence interval width (CIW) of the estimates $\\bar{Q}$. We present the subset of $Q$s with the worst performance in terms of bias: the mean and standard deviation of $Y$, and the regression coefficient of $X_1$. The solid gray line depicts the targets (unbiased estimates and nominal coverage rate)."}
mean_bias + sd_bias + Rsq_bias + est_bias + est_cov + est_ciw + plot_layout(ncol = 2, guides = "collect") + plot_annotation(tag_levels = "A", tag_suffix = ".")
```

>>>>

```{r Qs, fig.height = 6.8, fig.cap="Impact of non-convergence on statistical inferences. Depicted are the bias, coverage rate (CR) and confidence interval width (CIW) of the estimates $\\bar{Q}$. We present the subset of $Q$s with the worst performance in terms of bias: the mean and standard deviation of $Y$, and the regression coefficient of $X_1$. The solid gray line depicts the targets (unbiased estimates and nominal coverage rate)."}
mean_bias + sd_bias + Rsq_bias + est_bias + est_cov + est_ciw + plot_annotation(tag_levels = "A", tag_suffix = ".") + plot_layout(ncol = 2, guides = "collect") 
```
